# The headline number

```{r}
#| include: false
source("../R/simulate.R"); source("../R/metrics.R"); source("../R/theme_eval.R")
df <- simulate_predictions()
```

Here is a model. It was trained to flag a condition from a risk score, it has
been evaluated on a held-out set of `r nrow(df)` patients, and its performance
is the one number every such model is eventually reduced to:

```{r}
auroc(df$label, df$score)
```

An AUROC of about **0.87**. That is a good model — comfortably better than a
coin, the kind of number that clears a reviewer, anchors an abstract, and gets
carried into a slide deck as *the* result. If the story stopped here, it would
be a reassuring one.

## The average is hiding its variance

The patients in this dataset are not interchangeable. They fall into three
strata of very different size — think of Fitzpatrick skin types on a
dermatology model, where the darkest-skinned group is routinely under 5% of the
training images. Ask the same question inside each stratum:

```{r}
#| echo: false
by_group <- sapply(sort(unique(df$group)), function(g) {
  s <- df[df$group == g, ]
  auroc(s$label, s$score)
})
knitr::kable(
  data.frame(Group = names(by_group), AUROC = round(by_group, 3)),
  align = "lr"
)
```

The single number was 0.87. Inside the smallest group it is **`r round(by_group["V-VI"], 2)`** — a different
model, in effect, for the patients who were least present when it was built.
The pooled figure is not wrong; it is dominated. The largest stratum sets the
average, and the average then speaks confidently on behalf of a group it barely
represents.

```{r}
#| echo: false
#| fig-cap: "One pooled AUROC (dashed) against the three subgroup values it averages over. The gap widens as the group shrinks."
d <- data.frame(group = names(by_group), auroc = by_group,
                n = as.integer(table(df$group)[names(by_group)]))
ggplot(d, aes(x = auroc, y = reorder(group, auroc))) +
  geom_vline(xintercept = auroc(df$label, df$score),
             linetype = "dashed", colour = EVAL_GREY) +
  geom_segment(aes(x = auroc(df$label, df$score), xend = auroc,
                   yend = reorder(group, auroc)), colour = EVAL_GREY) +
  geom_point(aes(colour = group, size = n)) +
  scale_colour_manual(values = group_cols, guide = "none") +
  scale_size_continuous(range = c(3, 7), guide = "none") +
  labs(title = "The headline number is the largest group's number",
       subtitle = "Dashed line: the single pooled AUROC",
       x = "AUROC", y = NULL) +
  theme_eval()
```

## What this book does about it

Splitting the AUROC by subgroup is the first move, and the rest of this book is
the argument that it is not nearly enough. A subgroup estimate is still just an
estimate, and in a stratum of a few dozen events it can be wildly imprecise
without looking it (Chapter 5). Discrimination — can the model *rank* patients
— is only one of three questions, and the other two, whether its probabilities
are *calibrated* (Chapter 6) and whether a single decision *threshold* serves
every group (Chapter 7), are where transported models usually fail.

To make any of that trustworthy, we need an evaluation harness we can see all
the way through. That is Chapters 3 and 4: a simulator whose answer we know, and
an AUROC we wrote ourselves and can prove is right. First, though, we should be
clear about what the three questions even are.
