# The data, and its known answer

```{r}
#| include: false
source("../R/simulate.R"); source("../R/metrics.R"); source("../R/theme_eval.R")
df <- simulate_predictions()
tt <- truth_table()
```

To trust a measuring instrument, you calibrate it against something whose
length you already know. Real evaluation data can never be that reference,
because with real data the truth (the model's actual discrimination in each
subgroup) is precisely what you are trying to estimate and never get to see.
So we simulate, and we simulate in a way that hands us the answer up front.

## A generator with a closed-form truth

The design is deliberately plain. Within each stratum, non-events get a latent
score drawn from a standard normal, and events get one shifted up by a
separation `d`:

```{r}
simulate_predictions
```

The move that makes the whole book possible is this: for two normals of equal
variance separated by `d`, the probability that a draw from the event
distribution exceeds a draw from the non-event distribution *is* the AUROC,
and it has a closed form.

$$
\text{AUROC} \;=\; P(X_{\text{event}} > X_{\text{non-event}})
            \;=\; \Phi\!\left(\frac{d}{\sqrt{2}}\right)
$$

```{r}
true_auroc
```

So we do not have to wonder whether an estimate is good. We can subtract it
from the right answer.

## The three strata

The separations and sizes are chosen to caricature a common situation: a model
that is sharp in the group that dominates the data and dull in the group that
does not. The small group is also the one where the signal is genuinely
weaker.

```{r}
#| echo: false
knitr::kable(tt, digits = 3,
             col.names = c("Group", "n", "Separation d", "True AUROC"))
```

Two things are held fixed on purpose. The prevalence is the same **`r PREVALENCE`**
in every stratum, so that no performance gap can later be blamed on a
difference in base rates. The gap will be about separation and sample size,
nothing else. And the seed is fixed, so every number in this book is
reproducible to the digit.

```{r}
#| echo: false
#| fig-cap: "Latent score distributions by outcome, within each stratum. As the separation d shrinks, the two distributions overlap more, and the model's job gets harder."
ggplot(df, aes(x = latent, fill = factor(label))) +
  geom_density(alpha = 0.55, colour = NA) +
  facet_wrap(~ factor(group, levels = c("I-II", "III-IV", "V-VI")), ncol = 1) +
  scale_fill_manual(values = c("0" = EVAL_GREY, "1" = EVAL_ACCENT),
                    labels = c("non-event", "event"), name = NULL) +
  labs(title = "The same model is three different problems",
       subtitle = "More overlap means less separable means a lower true AUROC",
       x = "Latent score", y = NULL) +
  theme_eval() +
  theme(axis.text.y = element_blank())
```

In `I-II` the two distributions barely touch; in `V-VI` they sit almost on top
of each other. That overlap is the true AUROC of `r round(tt$true_auroc[tt$group == "V-VI"], 3)`, drawn.

The `score` column the rest of the book uses is just this latent value squeezed
through `plogis()` onto the 0–1 probability scale: a stand-in for a model that
is imperfectly calibrated, which becomes the subject of Chapter 6. Everything
downstream reads only `label` and `score`, exactly as it would for a real
model's output.

With a world whose answer we know, we can now write an estimator and prove it
recovers that answer.
