# Discrimination from scratch

```{r}
#| include: false
source("../R/simulate.R"); source("../R/metrics.R"); source("../R/theme_eval.R")
df <- simulate_predictions()
tt <- truth_table()
```

The AUROC is usually drawn as the area under a curve of sensitivity against
one-minus-specificity, swept over every threshold. That picture is fine, but it
hides how simple the quantity is. The area under the ROC curve is *exactly* the
probability that a randomly chosen event scores higher than a randomly chosen
non-event. Stated that way, it is a rank statistic, and it can be computed in
one line.

## The whole estimator

```{r}
auroc
```

Rank all the scores; sum the ranks belonging to the events; subtract the ranks
those events would occupy if they were all crammed at the bottom
($n_1(n_1+1)/2$); divide by the number of event/non-event pairs $n_1 n_0$. What
comes out is the fraction of pairs the model orders correctly: the
Mann-Whitney U statistic, rescaled. Ties contribute exactly one half, and using
average ranks builds that half-credit in with no special case.

## Is it right? Two checks

A one-line estimator is easy to get subtly wrong, so we never take it on faith.
First, against a completely different code path (`wilcox.test`, which computes
the same U statistic through R's own rank machinery):

```{r}
w <- wilcox.test(score ~ factor(label), data = df, exact = FALSE)
u_ref <- unname(w$statistic) / (sum(df$label == 0) * sum(df$label == 1))
c(ours = auroc(df$label, df$score), wilcox = 1 - u_ref)
```

Identical to machine precision. That pins the *implementation*. But a
numerically perfect estimator of the wrong quantity is still wrong. The
second check is against the truth the simulator planted, the analytic
$\Phi(d/\sqrt{2})$ from the previous chapter:

```{r}
#| echo: false
est <- sapply(tt$group, function(g) {
  s <- df[df$group == g, ]; auroc(s$label, s$score)
})
knitr::kable(
  data.frame(Group = tt$group, `True AUROC` = round(tt$true_auroc, 3),
             Estimate = round(est, 3),
             Error = round(est - tt$true_auroc, 3), check.names = FALSE),
  align = "lrrr"
)
```

Close everywhere, and closest where the sample is largest. That is exactly the
behaviour a consistent estimator should show. The estimate in `V-VI` is
**`r round(est["V-VI"], 3)`** against a truth of `r round(tt$true_auroc[tt$group == "V-VI"], 3)`: an error of
`r round(est["V-VI"] - tt$true_auroc[tt$group == "V-VI"], 3)` on 160 patients. Whether that gap is
alarming or just sampling noise is not a question the point estimate can
answer, and it is the subject of the next chapter.

## What discrimination cannot see

Because it is built from ranks, the AUROC is invariant to any strictly
increasing transformation of the score. Running it on the raw latent value and
on the squeezed probability gives the same answer to the last digit:

```{r}
c(on_probability = auroc(df$label, df$score),
  on_latent      = auroc(df$label, df$latent))
```

This is a feature and a warning at once. A feature, because it means a badly
scaled or miscalibrated model can still discriminate perfectly, and ranking is
the one property recalibration cannot destroy. A warning, because it means the
AUROC is structurally blind to whether the probabilities mean anything at all.
A model can post an AUROC of `r round(auroc(df$label, df$score), 2)` and still be uninterpretable and
unsafe at any given threshold.

AUROC sees the order and nothing else. To find out whether the numbers
themselves can be trusted (and whether the `r round(est["V-VI"], 3)` in `V-VI` is real), we need to
put a confidence interval on it, and then a second question entirely.
