# Calibration, and a metric that lies

```{r}
#| include: false
source("../R/simulate.R"); source("../R/metrics.R"); source("../R/theme_eval.R")
df <- simulate_predictions()
rep <- stratified_report(df, n_boot = 500L)
```

Chapter 4 established that AUROC cannot see the numbers, only their order.
This chapter asks the question it cannot answer: when the model says 0.8,
does the event happen 80% of the time?

## Two ways to measure it

**The calibration slope** regresses the outcome on the logit of the predicted
probability. Perfect calibration gives slope 1, intercept 0. Slope below 1
means predictions are too extreme; above 1, too conservative.

```{r}
calibration
```

**Expected calibration error** bins the predictions and averages the absolute
gap between predicted and observed rates within each bin. It is the metric
most commonly reported in the machine-learning literature.

```{r}
ece
```

Both are reasonable. On this data they disagree completely.

## The disagreement

```{r}
#| echo: false
knitr::kable(rep[, c("group","cal_slope","ece","brier","auroc")], digits = 3,
             col.names = c("Group","Calibration slope","ECE","Brier","AUROC"))
```

The slope says `V-VI` is badly miscalibrated: 0.50 against 1.85 in `I-II`.
These are opposite failures: predictions too extreme in one group, too
conservative in the other. Pooled, they average to 1.60, a number describing
no one in the dataset.

ECE says all three groups are the same, to within 0.06.

## Why ECE misses it

ECE takes an absolute value inside each bin. Systematic over-prediction in
one part of the score range and under-prediction in another therefore
contribute in the same direction, and the tilt in the whole relationship
never shows up. It measures *how far off* predictions are, not
*in which direction, systematically*.

The slope, being a regression coefficient, sees the tilt directly.

```{r}
#| echo: false
#| fig-cap: "Observed event rate against predicted probability, by stratum. The diagonal is perfect calibration."
bins <- cut(df$score, breaks = seq(0, 1, 0.1), include.lowest = TRUE)
cal <- aggregate(cbind(obs = label, pred = score) ~ bins + group, data = df, FUN = mean)
n_bin <- aggregate(label ~ bins + group, data = df, FUN = length)
cal$n <- n_bin$label
ggplot(cal[cal$n >= 5, ], aes(pred, obs, colour = group)) +
  geom_abline(linetype = "dashed", colour = EVAL_GREY) +
  geom_line(linewidth = 0.7) +
  geom_point(aes(size = n)) +
  scale_colour_manual(values = group_cols) +
  scale_size_continuous(range = c(1.2, 4.5), guide = "none") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(title = "One line is flat, and one metric didn't notice",
       subtitle = "Point size is bin count",
       x = "Predicted probability", y = "Observed event rate") +
  theme_eval()
```

The `V-VI` line is visibly flatter than the diagonal: as the model's
confidence rises, the observed rate barely follows. That is a slope of 0.5,
drawn.

## Why this matters more than discrimination

If the model feeds a ranked worklist, miscalibration is survivable, because
order is preserved. If it feeds a **threshold**, it is not. A referral rule of
"investigate if p > 0.7" assumes 0.7 means the same thing in every patient.
Here it does not, and the group where it means least is the group already
least likely to have been examined by a specialist.

This is the standard signature of a transported model, and it is why
calibration is the first thing to break and the last thing to be reported
[@vancalster2019]. TRIPOD+AI now asks for calibration to be reported
explicitly, with a plot, for exactly this reason [@collins2024].

::: {.callout-note}
## Report both, and the plot

The slope and intercept catch systematic tilt; ECE catches local wobble; the
plot catches things neither summarises. Anyone reporting a single calibration
number is choosing which failures they are willing to miss.
:::
