# Thresholds, and the cost of one rule

```{r}
#| include: false
source("../R/simulate.R"); source("../R/metrics.R"); source("../R/theme_eval.R")
df  <- simulate_predictions()
eo  <- equalised_odds(df, threshold = 0.5)
gt  <- group_thresholds(df, target = 0.90)
```

Discrimination and calibration are properties of the scores. A decision is not:
it needs a threshold, a line past which the model says *act*. The moment you
draw that line, a new question opens: is one line right for everyone? Whenever
the subgroups differ in separability, the honest answer is no.

## One threshold, measured two ways

Start with the naive policy: the same cutoff, 0.5, for every group. The
equalised-odds view reports the true-positive and false-positive rates each
group actually gets under that shared rule [@hardt2016].

```{r}
equalised_odds
```

```{r}
#| echo: false
knitr::kable(
  data.frame(Group = names(eo$tpr),
             TPR = round(eo$tpr, 3),
             FPR = round(eo$fpr, 3)),
  align = "lrr"
)
```

Read the two columns separately, because they tell opposite stories. The
false-positive rates are close: the non-event scores are drawn from the same
distribution in every stratum, so a fixed cutoff catches roughly the same
fraction of them everywhere. It is the **true-positive** rate that collapses:
a gap of `r round(eo$tpr_gap, 3)` between the best and worst group. A shared threshold does not
share the harm evenly; it hands the worst-separated group a much lower chance
of being caught. That is the disparate impact hiding inside a group-blind rule.

## Fixing the clinical standard instead

So flip it. Fix the thing a clinician actually cares about ("detect 90% of
events") and let each group have whatever threshold reaches it. Now the
sensitivity is equal by construction, and the cost surfaces somewhere else.

```{r}
group_thresholds
```

```{r}
#| echo: false
knitr::kable(gt, digits = 3,
             col.names = c("Group", "Threshold", "Sensitivity", "Specificity"))
```

Every group now catches 90% of events. But look at what that costs in
specificity. In `I-II` it is `r round(gt$specificity[gt$group == "I-II"], 3)`, so you wave through most true
negatives. In `V-VI` it is `r round(gt$specificity[gt$group == "V-VI"], 3)`: to reach the same
sensitivity on a flatter ROC curve, the threshold slides down into the
non-event mass and specificity falls off a cliff.

Put in the terms a service feels: at the same clinical standard, the
best-served group refers about **`r round((1 - gt$specificity[gt$group == "I-II"]) * 100)`% of its
negatives** for follow-up, and the worst-served group refers about
**`r round((1 - gt$specificity[gt$group == "V-VI"]) * 100)`%** of its negatives. Same model,
same rule, wildly different burden. That is a service-design problem, not a
rounding error.

## The trade-off is forced, not incidental

You cannot escape by being cleverer with the cutoff. Equalised odds asks for
equal TPR *and* equal FPR at once; when two groups sit on ROC curves of
different quality, you can satisfy that only where the curves happen to cross,
which is nowhere near a clinically useful operating point. A shared threshold
fails on the TPR leg; equalising sensitivity fails on the FPR leg. Every choice
absorbs the disparity somewhere.

```{r}
#| echo: false
#| fig-cap: "The two policies, drawn. Left: one threshold gives equal FPR but unequal sensitivity. Right: equal 90% sensitivity gives sharply unequal specificity. Neither is free."
library(ggplot2)
d1 <- data.frame(group = names(eo$tpr), value = eo$tpr, panel = "Shared 0.5 threshold\n(sensitivity diverges)")
d2 <- data.frame(group = gt$group, value = gt$specificity, panel = "Equal 90% sensitivity\n(specificity diverges)")
d1$metric <- "Sensitivity"; d2$metric <- "Specificity"
dd <- rbind(d1, d2)
ggplot(dd, aes(x = value, y = factor(group, levels = c("V-VI","III-IV","I-II")),
               colour = group)) +
  geom_segment(aes(x = 0, xend = value,
                   yend = factor(group, levels = c("V-VI","III-IV","I-II"))),
               colour = EVAL_GREY) +
  geom_point(size = 3.6) +
  facet_wrap(~ panel) +
  scale_colour_manual(values = group_cols, guide = "none") +
  xlim(0, 1) +
  labs(title = "You choose which disparity to absorb",
       x = NULL, y = NULL) +
  theme_eval()
```

Which failure to accept (more missed cases in the worst group, or more false
alarms there) is a value judgement about who bears the cost of the model's
weaker grip on that group. It is not a judgement a metric can make for you, and
it is certainly not one a single shared threshold makes *honestly*; it just
makes it silently.

::: {.callout-important}
## A threshold is a policy, not a default

`p > 0.5` looks neutral and is not. It encodes equal false-positive rates and
unequal sensitivities, and it hides that choice behind a round number. Report
the operating point you actually use, the sensitivity and specificity it
produces *in each subgroup*, and who pays for the gap.
:::
